---
title: "two-state modeling"
output: html_document
date: "2026-04-26"
---


This R markdown file contains code to run two-state modeling on the combinatorial SCR library tested in the first SCR Macrophage manuscript. 
I recommend running this as code chunks in Rstudio. 


First: Define 2 paths: 
  1. Directory where you have downloaded `phago_suma_combined.Rds`
  2. Directory where you want to save plots from running this script. If using Rstudio, the plots will be printed to in-line. You can comment out the ggsave lines if you do not want to save these plots. 
```{r}
data_folder <- "/Users/jodielunger/OneDrive - Stanford/Code/combo_lib_modeling/"
plot_folder <- "/Users/jodielunger/OneDrive - Stanford/Code/combo_lib_modeling/"
```


# ===================================
# Steps for running two-state model
# ===================================


# Step 1 - Load Required Packages
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(stringr)
library(forcats)
library(tibble)
library(rsample)
library(ggh4x)
```

# Step 2 - Load Data
The measurements used are:
- MFI values as raw flow cytometry readouts from surface marker analysis
- Phagocytosis index: Raw "red total intensity" readout from the Incucyte as a proxy for phagocytosis of pHrodo-labeled E. coli

```{r}
phenotype_data <- readRDS(file.path(data_folder, "phago_suma_combined.Rds"))
head(phenotype_data)
```

Filter data to only include samples which were transduced with an SCR
```{r}
synth_df <- subset(phenotype_data, transduction_status == "T")
```


# Step 3 - Create definitions for analysis
```{r}
# Define phenotype columns present in your data
phenotype_cols   <- c("CD163_MFI","CD80_MFI","CD206_MFI","CD40_MFI","PDL1_MFI", "phagocytosis_index")
motif_cols <- c("motif_ID_1", "motif_ID_2", "motif_ID_3")

# Benchmarks (keep in clustering)
bench_samples <- c("GM-CSF","IFNγ","IFNγ + LPS","IL-10","IL-4","UT","empty")
```

# Step 4 - Count how many of each motif is in each SCR
```{r}
# Add a unique row_id to preserve sample identity across reshaping
# operations (pivot_longer / pivot_wider).
orig_df <- synth_df %>%
  mutate(row_id = row_number())

# ---- Convert motif columns to long format ----
# This creates one row per (sample, motif copy), allowing motifs
# to be counted if they appear multiple times within a sample.
motifs_long <- orig_df %>%
  pivot_longer(
    cols = all_of(motif_cols),   # melt motif_ID_1/2/3 into a single column
    names_to = "slot",           # original position (not used downstream)
    values_to = "motif"          # motif identity (e.g., "S1", "S2")
  ) %>%
  # Remove empty motif entries 
  # (samples with fewer than 3 motifs will have "" in places where the 1st, 2nd, or 3rd motif would go)
  filter(!is.na(motif), motif != "") %>%
  # Group by sample and motif to count how many times
  # each motif appears within each construct
  group_by(row_id, motif) %>%
  summarise(
    cnt = n(),                   # number of copies of this motif in the sample
    .groups = "drop"
  )

# ---- Pivot to wide format to create a motif count matrix ----
# Each motif becomes its own column
# and each cell contains the copy number of that motif in the sample.
motifs_wide <- motifs_long %>%
  pivot_wider(
    names_from  = motif,         # motif identities become column names
    values_from = cnt,           # fill cells with motif copy counts
    values_fill = 0              # motifs not present get a count of 0
  )

# This is an important line that ensures motifs all get labeled as having zero copies in SCR0
motif_count_cols <- setdiff(names(motifs_wide), "row_id") # list the motif columns so that NAs are replaced as 0 ONLY in these columns!!

# ---- Join motif counts back to the original dataframe ----
# This produces the final modeling dataframe, where:
#   - original sample metadata and phenotypes are preserved
#   - each motif column encodes copy number (0, 1, 2, ...)
synth_df <- orig_df %>%
  left_join(motifs_wide, by = "row_id") %>%
  mutate(across(all_of(motif_count_cols), ~ tidyr::replace_na(.x, 0))) %>%
  select(-row_id) # remove temporary identifier 
```



# Step 5 - Summarize replicates
Create version of synth_df with averaged replicates
```{r}
# ------------------------------------------------------------------------------
# Aggregate Replicates and Calculate Standard Errors
# ------------------------------------------------------------------------------
synth_df_aggregated <- synth_df %>%
  group_by(sample_name) %>%
  summarise(
    # ------------------------------------------------------------
    # Keep motif counts as columns (identical across replicates)
    # ------------------------------------------------------------
    across(all_of(c("S1", "S2", "S3", "S4", "S5", "S6", "S7", "S8", "S9")), dplyr::first),
    
    
    # ------------------------------------------------------------
    # Raw measurement columns (from your phenotype_cols vector): mean, SE, and n
    # -> produces columns like CD40_MFI_mean, CD40_MFI_se, CD40_MFI_n
    # ------------------------------------------------------------
    across(
      all_of(phenotype_cols),
      list(
        mean = ~ mean(., na.rm = TRUE),
        se   = ~ sd(., na.rm = TRUE) / sqrt(sum(!is.na(.))),
        n    = ~ sum(!is.na(.))
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  ) 
```

# Step 6 - Plot phenotype distributions and store extreme values

```{r}
# ------------------------------------------------------------------------------
# Determine f_low and f_high from MFI and phagocytosis data (all phenotype data)
# ------------------------------------------------------------------------------

# Assign mean columns
phenotype_mean_cols <- synth_df_aggregated %>%
  select(ends_with("_mean")) %>%
  colnames()


# Helper: estimate two-state expression bounds for one phenotype column
# f_low  = lowest observed value 
# f_high = highest observed value

# Function to plot distributions of each phenotype and store the extreme (min and max) values 
estimate_two_state_extremes <- function(df, pheno_col, k = 1, make_plot = TRUE) {
  stopifnot(pheno_col %in% names(df))

  d <- df %>%
    select(sample_name, S1:S9, all_of(pheno_col)) %>%
    drop_na(all_of(pheno_col))

  if (nrow(d) == 0) {
    return(list(
      results = tibble(
        phenotype = pheno_col,
        n_samples = 0L,
        f_low = NA_real_,
        f_high = NA_real_
      ),
      plot = NULL,
      sorted = tibble()
    ))
  }

  # Sort like your phago code
  sorted <- d %>%
    arrange(.data[[pheno_col]]) %>%
    select(sample_name, !!pheno_col)

  k_use <- min(k, nrow(sorted))

  low_samples  <- head(sorted$sample_name, k_use)
  high_samples <- tail(sorted$sample_name, k_use)

  f_low  <- mean(head(sorted[[pheno_col]], k_use), na.rm = TRUE)
  f_high <- mean(tail(sorted[[pheno_col]], k_use), na.rm = TRUE)

  # Mark which samples were used
  d_plot <- d %>%
    mutate(
      bound_group = dplyr::case_when(
        sample_name %in% low_samples  ~ paste0("Bottom ", k_use, " (f_low)"),
        sample_name %in% high_samples ~ paste0("Top ", k_use, " (f_high)"),
        TRUE                          ~ "Other"
      ),
      bound_group = factor(
        bound_group,
        levels = c(paste0("Bottom ", k_use, " (f_low)"),
                   paste0("Top ", k_use, " (f_high)"),
                   "Other")
      )
    )

  p <- NULL
  if (isTRUE(make_plot)) {
    p <- ggplot(d_plot, aes(x = .data[[pheno_col]])) +
      geom_histogram(bins = 30, fill = "grey85", color = "grey30", linewidth = 0.3) +
      # Color ONLY the selected samples on the x-axis
      geom_rug(
        data = d_plot %>% filter(bound_group != "Other"),
        aes(color = bound_group),
        sides = "b",
        alpha = 0.95,
        linewidth = 1.0
      ) +
      # Dotted lines at the averages used for bounds
      geom_vline(xintercept = f_low,  linetype = "dotted", linewidth = 0.8) +
      geom_vline(xintercept = f_high, linetype = "dotted", linewidth = 0.8) +
      labs(
        title = paste0("Phenotype Distribution: ", pheno_col),
        subtitle = paste0(
          "Colored ticks = samples used for bounds; dotted lines = f_low (",
          round(f_low, 2), ") and f_high (", round(f_high, 2), ")"
        ),
        x = pheno_col,
        y = "Count",
        color = "Used for bounds"
      ) +
      theme_classic()
  }

  list(
    results = tibble(
      phenotype = pheno_col,
      n_samples = nrow(d),
      f_low = f_low,
      f_high = f_high
    ),
    plot = p,
    sorted = sorted
  )
}

# Run function across all phenotype_mean columns
phenotype_extremes_list <- purrr::map(
  phenotype_mean_cols,
  ~ estimate_two_state_extremes(
    df = synth_df_aggregated,
    pheno_col = .x,
    k = 1,
    make_plot = TRUE
  )
)

# Tidy summary table
phenotype_extremes_tbl <- purrr::map_dfr(phenotype_extremes_list, "results")
print(phenotype_extremes_tbl)

# Display all histograms in RMarkdown
purrr::walk(phenotype_extremes_list, ~ if (!is.null(.x$plot)) print(.x$plot))

```


# Step 7 - Fit the two-state model


The two-state model assumes cells exist as a population of cells in two states, where the fraction of cells in each state determines the resulting phenotype measurement for each phenotype.

$$
f_{obs} = f_{low} \cdot (1 - [ON]) + f_{high} \cdot [ON]
$$


Where:

- $f_{obs}$ = observed fluorescence value for the sample
- $f_{low}$ = lowest observed fluorescence value for that phenotype (across the whole dataset)
- $f_{high}$ = highest observed fluorescence value for that phenotype (across the whole dataset)
- $[ON]$ = fraction of cells in the ON (high) state
- $(1 - [ON])$ = fraction of cells in the OFF (low) state

The fraction of cells in the ON state is determined by the equilibrium constant $K$:

$$
[ON] = \frac{K}{1 + K}
$$

where the equilibrium constant relates to a change in energy between the low and high phenotype state:

$$
K = e^{-\Delta E}
$$


Substituting this into the fraction equation:

$$
[ON] = \frac{e^{-\Delta E}}{1 + e^{-\Delta E}} = \frac{1}{1 + e^{\Delta E}}
$$


Combining these relationships, the model becomes:

$$
f_{obs} = f_{low} \cdot \left(1 - \frac{1}{1 + e^{\Delta E}}\right) + f_{high} \cdot \frac{1}{1 + e^{\Delta E}}
$$

where $\Delta E$ follows the same additive motif model as for the surface markers:

$$
\Delta E = \Delta E_i + \Delta E_{s,S1} \cdot n_{S1} + \Delta E_{s,S2} \cdot n_{S2} + ... + \Delta E_{s,S9} \cdot n_{S9}
$$
where $\Delta E_{i}$ is the intrinsic $\Delta E$ value when there are no motifs present in the SCR, and $\Delta E_{s,Sx}$ is the per-motif (single-motif) $\Delta E$ contribution of motif $S_x$. The total $\Delta E$ is the sum of the intrinsic value and all per-motif contributions.



```{r}
motif_cols <- paste0("S", 1:9)

# Individual replicates used for model fitting
fit_df <- synth_df %>%
  select(sample_name, all_of(motif_cols), all_of(phenotype_cols)) %>%
  drop_na(all_of(motif_cols))
```


```{r}
safe_cor <- function(x, y) {
  if (length(x) < 3) return(NA_real_)
  if (sd(x, na.rm = TRUE) == 0 || sd(y, na.rm = TRUE) == 0) return(NA_real_)
  cor(x, y, use = "complete.obs")
}

rmse <- function(x, y) sqrt(mean((x - y)^2, na.rm = TRUE))
```

```{r}
# fit_two_state_model fits a two-state model in MFI space for one phenotype.
# Parameters (ΔEi, ΔEs_S1..S9) are estimated by minimizing MFI residuals on
# individual replicates via BFGS. r and R² are evaluated on averaged replicates.
# Coefficients are returned in the same format used for Ki/Ks plots and tables.

fit_two_state_model <- function(phenotype_mean_col, n_folds = 10, seed = 123) {

  eps               <- 1e-6
  phenotype_raw_col <- str_remove(phenotype_mean_col, "_mean$")
  term_names        <- c("(Intercept)", motif_cols)

  # ---- Step 1: Bounds ----
  ext    <- phenotype_extremes_tbl %>% filter(.data$phenotype == phenotype_mean_col) %>% slice(1)
  f_low  <- ext$f_low
  f_high <- ext$f_high

  # ---- Step 2: Build datasets ----
  # d_ind: individual replicates — fitting and CV
  # d_avg: averaged replicates  — predictions and r
  d_ind <- fit_df %>%
    select(sample_name, all_of(motif_cols), all_of(phenotype_raw_col)) %>%
    rename(f_obs = all_of(phenotype_raw_col)) %>%
    drop_na(f_obs) %>%
    mutate(
      frac_on = pmin(pmax((f_obs - f_low) / (f_high - f_low), eps), 1 - eps),
      dE_obs  = log((1 - frac_on) / frac_on)
    )

  d_avg <- synth_df_aggregated %>%
    select(sample_name, all_of(motif_cols), all_of(phenotype_mean_col)) %>%
    rename(f_obs = all_of(phenotype_mean_col)) %>%
    drop_na(f_obs)

  cat(sprintf("\n%s | f_low=%.1f, f_high=%.1f | N_ind=%d, N_avg=%d\n",
              phenotype_mean_col, f_low, f_high, nrow(d_ind), nrow(d_avg)))

  mm_ind <- as.matrix(d_ind[, motif_cols])

  # Two-state MFI prediction: params = c(dEi, dEs_S1..S9)
  predict_mfi <- function(params, mm) {
    f_on <- 1 / (1 + exp(params[1] + as.vector(mm %*% params[-1])))
    f_low * (1 - f_on) + f_high * f_on
  }

  mfi_loss <- function(params, f_obs, mm)
    sum((f_obs - predict_mfi(params, mm))^2, na.rm = TRUE)

  # Initialize ΔEi from no-motif mean; motif ΔEs at 0
  no_motif    <- rowSums(mm_ind) == 0
  init_params <- c(
    if (any(no_motif)) mean(d_ind$dE_obs[no_motif], na.rm = TRUE) else 0,
    rep(0, length(motif_cols))
  )

  # ---- Step 3: Fit ----
  fit_all <- try(
    optim(init_params, mfi_loss, f_obs = d_ind$f_obs, mm = mm_ind,
          method = "BFGS", control = list(maxit = 5000)),
    silent = TRUE
  )

  if (inherits(fit_all, "try-error") || fit_all$convergence != 0)
    warning(sprintf("optim did not converge for %s (code %d)", phenotype_mean_col,
                    if (inherits(fit_all, "try-error")) -1L else fit_all$convergence))

  if (inherits(fit_all, "try-error"))
    return(list(phenotype = phenotype_mean_col, coef_tbl = tibble(), pred_tbl = tibble(),
                quality_tbl = tibble(phenotype = phenotype_mean_col, status = "model fit failed",
                                     n_samples = nrow(d_avg), n_ind = nrow(d_ind))))

  coefs_all <- tibble(term = term_names, estimate = fit_all$par)

  # ---- Step 4: k-fold CV ----
  set.seed(seed)
  folds <- vfold_cv(d_ind, v = n_folds)

  fold_results <- map(folds$splits, function(split) {
    train <- analysis(split)
    test  <- assessment(split)
    mm_tr <- as.matrix(train[, motif_cols])
    mm_te <- as.matrix(test[, motif_cols])

    fit_fold <- try(
      optim(fit_all$par, mfi_loss, f_obs = train$f_obs, mm = mm_tr,
            method = "BFGS", control = list(maxit = 2000)),
      silent = TRUE
    )

    if (inherits(fit_fold, "try-error"))
      return(list(coefs = tibble(term = term_names, estimate = NA_real_),
                  oof   = tibble(f_obs = test$f_obs, f_pred = NA_real_)))

    list(
      coefs = tibble(term = term_names, estimate = fit_fold$par),
      oof   = tibble(f_obs = test$f_obs, f_pred = predict_mfi(fit_fold$par, mm_te))
    )
  })

  cv_sd <- map_dfr(fold_results, "coefs", .id = "fold") %>%
    group_by(term) %>%
    summarise(std.dev = sd(estimate, na.rm = TRUE), .groups = "drop")

  oof    <- map_dfr(fold_results, "oof")
  cv_r2  <- safe_cor(oof$f_obs, oof$f_pred)^2

  coef_tbl <- coefs_all %>%
    left_join(cv_sd, by = "term") %>%
    mutate(phenotype = phenotype_mean_col, f_low = f_low, f_high = f_high)

  # ---- Step 5: Predict on averaged data ----
  mm_avg      <- as.matrix(d_avg[, motif_cols])
  f_pred      <- predict_mfi(fit_all$par, mm_avg)
  motif_count <- rowSums(mm_avg)

  pred_tbl <- tibble(
    phenotype   = phenotype_mean_col,
    sample_name = d_avg$sample_name,
    f_obs       = d_avg$f_obs,
    f_pred      = f_pred,
    motif_count = motif_count
  ) %>%
    mutate(
      motif_group = case_when(
        motif_count == 0 ~ "0 motifs",
        motif_count == 1 ~ "1 motif",
        TRUE             ~ "2+ motifs"
      ),
      motif_group = factor(motif_group, levels = c("0 motifs", "1 motif", "2+ motifs"))
    )

  # ---- Step 6: Quality metrics on averaged data ----
  r <- safe_cor(pred_tbl$f_obs, pred_tbl$f_pred)

  quality_tbl <- tibble(
    phenotype = phenotype_mean_col,
    r         = r,
    r2        = r^2,
    cv_r2     = cv_r2,
    rmse      = rmse(pred_tbl$f_obs, pred_tbl$f_pred),
    n_samples = nrow(d_avg),
    n_ind     = nrow(d_ind),
    f_low     = f_low,
    f_high    = f_high,
    status    = if (fit_all$convergence == 0) "ok" else paste0("convergence code ", fit_all$convergence)
  )

  list(phenotype = phenotype_mean_col, coef_tbl = coef_tbl, pred_tbl = pred_tbl, quality_tbl = quality_tbl)
}
```

Run model fitting across all phenotypes and inspect quality metrics
```{r}

# Fit the model for every phenotype
fit_list <- map(phenotype_mean_cols, fit_two_state_model)

# Stack output tables across all phenotypes
results_coef    <- map_dfr(fit_list, "coef_tbl")
results_pred    <- map_dfr(fit_list, "pred_tbl")
results_quality <- map_dfr(fit_list, "quality_tbl")

results_quality %>%
  select(phenotype, r, r2, cv_r2, rmse, n_samples, n_ind, status) %>%
  arrange(desc(r2))
```


Plot in-sample vs cross-validated R² per phenotype
```{r}
results_quality %>%
  select(phenotype, r2, cv_r2) %>%
  pivot_longer(
    cols      = c(r2, cv_r2),
    names_to  = "metric",
    values_to = "r2"
  ) %>%
  mutate(
    metric   = recode(metric,
                      r2    = "In-sample R²",
                      cv_r2 = "Cross-validated R²"),
    metric   = factor(metric, levels = c("In-sample R²", "Cross-validated R²")),
    phenotype = fct_reorder(phenotype, r2, .fun = max, .desc = TRUE)
  ) %>%
  ggplot(aes(x = phenotype, y = r2, fill = metric)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c("In-sample R²" = "steelblue", "Cross-validated R²" = "coral")) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  theme_classic() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    legend.title = element_blank()
  ) +
  labs(
    title = "Model fit: in-sample vs cross-validated R² by phenotype",
    x     = NULL,
    y     = "R²"
  )
```

# Step 8 - Plot observed vs. predicted
```{r}
# ==============================================================================
# Observed vs Predicted plots for each phenotype
# ==============================================================================

# Color/alpha scheme:
motif_colors <- c("0 motifs" = "black", "1 motif" = "palevioletred1", "2+ motifs" = "black")
motif_alphas <- c("0 motifs" = 1, "1 motif" = 1, "2+ motifs" = 0.12)

# Log transform predicted phenotype values 
# ==============================================================================
log_f_predictions <- results_pred %>%
  mutate(
    phenotype_label = str_remove(phenotype, "_mean$"),
    motif_group = factor(motif_group, levels = c("0 motifs", "1 motif", "2+ motifs")),
    # Log-transform phenotype (log 10)
    f_obs_log = log10(f_obs),
    f_pred_log = log10(f_pred)
  )

# Compute r and R² on raw values, display on log-transformed plot
# ==============================================================================
log_phenotype_annotations <- log_f_predictions %>%
  group_by(phenotype, phenotype_label) %>%
  summarise(
    r = suppressWarnings(cor(f_obs, f_pred, use = "complete.obs", method = "pearson")),
    r2 = r^2,
    x_pos = min(f_obs_log, na.rm = TRUE) +
      0.05 * (max(f_obs_log, na.rm = TRUE) - min(f_obs_log, na.rm = TRUE)),
    y_pos = max(f_pred_log, na.rm = TRUE) -
      0.05 * (max(f_pred_log, na.rm = TRUE) - min(f_pred_log, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    label = ifelse(is.finite(r),
                   sprintf("r = %.2f\nR² = %.2f", r, r2),
                   "r = NA\nR² = NA")
  )


# Get matched x/y limits per phenotype (for plotting)
# ==============================================================================
facet_limits <- log_f_predictions %>%
  group_by(phenotype_label) %>%
  summarise(
    min_lim = min(c(f_obs_log, f_pred_log), na.rm = TRUE),
    max_lim = max(c(f_obs_log, f_pred_log), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pad = 0.05 * (max_lim - min_lim),
    min_lim = min_lim - pad,
    max_lim = max_lim + pad
  )
x_scales <- lapply(seq_len(nrow(facet_limits)), function(i) {
  scale_x_continuous(limits = c(facet_limits$min_lim[i], facet_limits$max_lim[i]))
})
y_scales <- lapply(seq_len(nrow(facet_limits)), function(i) {
  scale_y_continuous(limits = c(facet_limits$min_lim[i], facet_limits$max_lim[i]))
})

# Plot observed vs. predicted
# ==============================================================================
p_phenotype_log_obs_pred <- ggplot(
  log_f_predictions,
  aes(x = f_obs_log, y = f_pred_log,
      color = motif_group, alpha = motif_group)
) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed",
              color = "red",
              alpha = 0.5,
              linewidth = 0.5) +
  geom_point(size = 2) +
  geom_text(
    data = log_phenotype_annotations,
    aes(x = x_pos, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 3,
    fontface = "bold",
    color = "black"
  ) +
  facet_wrap(~ phenotype_label, scales = "free") +
  facetted_pos_scales(
    x = x_scales,
    y = y_scales
  ) +
  scale_color_manual(values = motif_colors, name = "Motif Count") +
  scale_alpha_manual(values = motif_alphas, guide = "none") +
  labs(
    x = "log10(f_obs)",
    y = "log10(f_pred)",
    title = "Model: Observed vs Predicted (log-transformed)",
  ) +
  theme_classic(base_size = 11) +
  theme(
      aspect.ratio = 1,
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_blank(),
    legend.position = "bottom",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )

p_phenotype_log_obs_pred

ggsave(
  filename = file.path(plot_folder, "obs_vs_pred_log_phenotype.svg"),
  plot = p_phenotype_log_obs_pred,
  width = 8,
  height = 6,
  device = "svg"
)

```



# Step 9 - Plot Ks values

The equilibrium constant $K = e^{-\Delta E}$ uses the total $\Delta E$. Each motif contributes a per-motif (single-motif) value $K_{s} = e^{-\Delta E_{s}}$, and the total $K$ is the product of all per-motif $K_{s}$ values (times the intrinsic $K_i$). We plot $K_{s}$ per motif to visualize each motif's individual multiplicative contribution to the equilibrium between positive and negative states.

- Ks > 1: Motif shifts equilibrium toward positive state (corresponds to ΔEs < 0)
- Ks = 1: Motif has no effect on equilibrium (corresponds to ΔEs = 0)
- Ks < 1: Motif shifts equilibrium toward negative state (corresponds to ΔEs > 0)

```{r}
# Each fitted coefficient is a per-motif ΔEs value; Ks = exp(-ΔEs).
# Error bars show ± 1 SD of each coefficient across CV folds (stability).
# results_coef columns used: term, estimate (ΔEs), std.dev (CV-fold SD), phenotype.

# Fix x-axis order: intercept first, then motifs S1-S9
term_order <- c("(Intercept)", paste0("S", 1:9))

# log10(e) ≈ 0.4343 — scaling factor used to convert ΔE units to log10(K) units
log10e <- log10(exp(1))

log10Ks_plot_df <- results_coef %>%
  mutate(
    # Fix term factor levels so x-axis always reads (Intercept), S1, S2, ..., S9
    term = factor(term, levels = term_order),

    # Strip "_mean" suffix for cleaner facet labels
    phenotype_label = str_remove(phenotype, "_mean$"),

    # ---- Convert ΔEs coefficient -> log10(Ks) effect ----
    # Each motif coefficient is a per-motif ΔEs value; Ks = exp(-ΔEs)
    # Taking log10:  log10(Ks) = -ΔEs * log10(e)
    # A positive log10(Ks) means the motif shifts equilibrium toward the high state.
    # A negative log10(Ks) means the motif shifts equilibrium toward the low state.
    log10Ks_estimate = -estimate * log10e,

    # ---- Propagate CV-fold SD from ΔEs space to log10(Ks) space ----
    # std.dev is the SD of the ΔEs coefficient across the k CV fold models.
    # Since log10(Ks) = -ΔEs * log10(e) is a linear transform (Y = aX),
    # SD propagates as: SD(Y) = |a| * SD(X)
    # Here |a| = log10(e), so: SD(log10Ks) = log10(e) * SD(ΔEs)
    # This is NOT a classical SE — it reflects fold-to-fold coefficient variability.
    log10Ks_sd = std.dev * log10e,

    # Error bar bounds: center ± 1 CV-fold SD in log10(Ks) units
    ymin = log10Ks_estimate - log10Ks_sd,
    ymax = log10Ks_estimate + log10Ks_sd
  )

# ---- Plot ----
# Each bar = log10(Ks) effect of one motif (or intercept) on one phenotype
# Error bars = ± 1 SD across CV folds (coefficient stability)
# Red dashed line at 0 = Ks of 1, meaning no shift in equilibrium
p_log10Ks <- ggplot(log10Ks_plot_df, aes(x = term, y = log10Ks_estimate)) +
  geom_col(width = 0.75) +
  geom_errorbar(
    aes(ymin = ymin, ymax = ymax),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed", color = "red") +
  facet_wrap(~ phenotype_label, scales = "free_y") +
  labs(
    x = NULL,
    y = "log10(Ks) effect (± CV fold SD)",
    title = "Per-motif equilibrium constant (Ks) effects",
    subtitle = "log10(Ks) > 0 shifts equilibrium toward positive state; log10(Ks) < 0 shifts toward negative state"
  ) +
  coord_cartesian(clip = "off") +
  theme_classic(base_size = 11) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.y = element_text(face = "bold"),
    panel.spacing = unit(0.9, "lines"),
    plot.margin = margin(8, 10, 8, 8)
  )

p_log10Ks

ggsave(
  filename = file.path(plot_folder, "motif_log10Ks_effects.svg"),
  plot = p_log10Ks,
  height = 5,
  width = 8,
  device = "svg"
)

```

# ===================================
# Export fitted values 

```{r}
coef_export_tbl <- results_coef %>%
  mutate(
    K_value    = exp(-estimate),
    term_label = ifelse(term == "(Intercept)", "Ki", paste0("Ks_", term)),
    dE_label   = ifelse(term == "(Intercept)", "dEi", paste0("dEs_", term)),
    phenotype  = str_remove(phenotype, "_mean$")
  ) %>%
  select(phenotype, term_label, dE_label, K_value, estimate) %>%
  pivot_wider(
    id_cols     = phenotype,
    names_from  = c(term_label, dE_label),
    values_from = c(K_value, estimate),
    names_glue  = "{term_label}_or_{dE_label}_{.value}"
  )

# Build tidy wide table: one row per phenotype, columns for K and ΔE values side by side
term_order_export <- c("(Intercept)", paste0("S", 1:9))
coef_export_tbl <- results_coef %>%
  mutate(phenotype = str_remove(phenotype, "_mean$")) %>%
  arrange(phenotype, match(term, term_order_export)) %>%
  mutate(
    K_value    = exp(-estimate),
    term_label = ifelse(term == "(Intercept)", "Ki",    paste0("Ks_", term)),
    dE_label   = ifelse(term == "(Intercept)", "dEi",   paste0("dEs_", term))
  ) %>%
  select(phenotype, term_label, dE_label, K_value, estimate) %>%
  pivot_longer(
    cols      = c(K_value, estimate),
    names_to  = "value_type",
    values_to = "value"
  ) %>%
  mutate(col_name = ifelse(value_type == "K_value", term_label, dE_label)) %>%
  select(phenotype, col_name, value) %>%
  pivot_wider(names_from = col_name, values_from = value) %>%
  select(phenotype,
         Ki,   dEi,
         Ks_S1, dEs_S1, Ks_S2, dEs_S2, Ks_S3, dEs_S3,
         Ks_S4, dEs_S4, Ks_S5, dEs_S5, Ks_S6, dEs_S6,
         Ks_S7, dEs_S7, Ks_S8, dEs_S8, Ks_S9, dEs_S9)

knitr::kable(coef_export_tbl, digits = 4,
             caption = "Fitted K and ΔE values per phenotype (Ki/dEi = intrinsic; Ks/dEs = per-motif)")

write.csv(coef_export_tbl, file = file.path(plot_folder, "Ki_Ks_dEi_dEs_table.csv"), row.names = FALSE)
```


# Predictions for all 0-4 motif SCRs 


Generate predictions and compute inflammatory score.
```{r}
# All (n_S1,...,n_S9) with each n_k >= 0 and total <= 4 (715 combinations)
all_combos <- expand.grid(setNames(rep(list(0:4), 9), paste0("S", 1:9))) %>%
  filter(rowSums(.) <= 4) %>%
  mutate(across(everything(), as.integer)) %>%
  as_tibble()

# Apply fitted model coefficients to predict each phenotype
predict_phenotype <- function(combos, phenotype_col, coef_data) {
  p      <- coef_data %>% filter(phenotype == phenotype_col)
  dEi    <- p$estimate[p$term == "(Intercept)"]
  dEs    <- p %>% filter(term != "(Intercept)") %>%
    arrange(match(term, paste0("S", 1:9))) %>%
    pull(estimate)
  f_low  <- p$f_low[1]
  f_high <- p$f_high[1]
  mm     <- as.matrix(combos[, paste0("S", 1:9)])
  f_on   <- 1 / (1 + exp(dEi + as.vector(mm %*% dEs)))
  f_low * (1 - f_on) + f_high * f_on
}

combo_predictions <- all_combos
for (pheno in phenotype_mean_cols) {
  pred_col <- paste0(str_remove(pheno, "_mean$"), "_pred")
  combo_predictions[[pred_col]] <- predict_phenotype(all_combos, pheno, results_coef)
}

# Normalize CD40 and CD80 predictions to 0-1 and average into an inflammatory score
combo_predictions <- combo_predictions %>%
  mutate(
    CD40_norm          = (CD40_MFI_pred - min(CD40_MFI_pred)) / (max(CD40_MFI_pred) - min(CD40_MFI_pred)),
    CD80_norm          = (CD80_MFI_pred - min(CD80_MFI_pred)) / (max(CD80_MFI_pred) - min(CD80_MFI_pred)),
    inflammatory_score = (CD40_norm + CD80_norm) / 2,
    total_motifs       = rowSums(across(all_of(paste0("S", 1:9)))),
    total_motifs_label = factor(
      case_when(
        total_motifs == 0 ~ "0 motifs",
        total_motifs == 1 ~ "1 motif",
        total_motifs == 2 ~ "2 motifs",
        total_motifs == 3 ~ "3 motifs",
        total_motifs == 4 ~ "4 motifs"
      ),
      levels = c("0 motifs", "1 motif", "2 motifs", "3 motifs", "4 motifs")
    ),
    combo_label = paste0(
      ifelse(S1 > 0, paste0("S1×", S1, " "), ""),
      ifelse(S2 > 0, paste0("S2×", S2, " "), ""),
      ifelse(S3 > 0, paste0("S3×", S3, " "), ""),
      ifelse(S4 > 0, paste0("S4×", S4, " "), ""),
      ifelse(S5 > 0, paste0("S5×", S5, " "), ""),
      ifelse(S6 > 0, paste0("S6×", S6, " "), ""),
      ifelse(S7 > 0, paste0("S7×", S7, " "), ""),
      ifelse(S8 > 0, paste0("S8×", S8, " "), ""),
      ifelse(S9 > 0, paste0("S9×", S9, " "), "")
    ) %>% str_trim() %>% ifelse(. == "", "No motif", .)
  )

# Flag combinations that were experimentally tested in the fitted data
tested_combos <- synth_df %>%
  mutate(across(all_of(paste0("S", 1:9)), as.integer)) %>%
  select(all_of(paste0("S", 1:9))) %>%
  distinct() %>%
  mutate(experimentally_tested = TRUE)

combo_predictions <- combo_predictions %>%
  left_join(tested_combos, by = paste0("S", 1:9)) %>%
  mutate(experimentally_tested = replace_na(experimentally_tested, FALSE))

# Pareto front: combinations that maximize both phagocytosis and inflammatory score
pareto_front <- combo_predictions %>%
  arrange(desc(phagocytosis_index_pred)) %>%
  mutate(
    best_inflam = cummax(inflammatory_score),
    is_pareto   = inflammatory_score >= best_inflam - 1e-12
  ) %>%
  filter(is_pareto) %>%
  arrange(phagocytosis_index_pred)

# Three special designs to highlight as colored squares
special_designs <- tibble(
  S1 = c(0L, 2L, 0L),
  S2 = c(0L, 0L, 0L),
  S3 = c(0L, 0L, 2L),
  S4 = c(0L, 0L, 0L),
  S5 = c(0L, 0L, 0L),
  S6 = c(0L, 0L, 0L),
  S7 = c(0L, 0L, 0L),
  S8 = c(0L, 0L, 0L),
  S9 = c(0L, 0L, 0L),
  fill_color = c("black", "#dea152", "#8e64a2")
) %>%
  left_join(combo_predictions, by = paste0("S", 1:9))
```




Plot to predict phagocytosis by inflammatory score for all motif combinations (0-4 motifs)
```{r}
p_design_space <- ggplot(combo_predictions,
                         aes(x = phagocytosis_index_pred, y = inflammatory_score)) +
  geom_point(data = filter(combo_predictions, !experimentally_tested),
             color = "grey80", size = 1.5, alpha = 0.4) +
  geom_point(data = filter(combo_predictions, experimentally_tested),
             color = "steelblue", size = 2, alpha = 0.9) +
  geom_line(data  = pareto_front,
            color = "black", linewidth = 0.8, linetype = "dashed") +
  geom_point(data = pareto_front, size = 2.5) +
  geom_text_repel(data = pareto_front, aes(label = combo_label),
                  size = 3, color = "black", max.overlaps = 20, box.padding = 0.5) +
  geom_point(data  = special_designs,
             aes(fill = fill_color),
             shape = 22, size = 5, stroke = 0) +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, 1), expand = c(0.01, 0.01)) +
  theme_classic(base_size = 11) +
  theme(panel.border    = element_rect(color = "black", fill = NA, linewidth = 0.5),
        legend.position = "none") +
  labs(x = "Predicted Phagocytosis Index",
       y = "Inflammatory Score\n(normalized CD40 + CD80) / 2")

p_design_space

ggsave(
  filename = file.path(plot_folder, "zero_to_four_motif_predictions.svg"),
  plot     = p_design_space,
  width    = 7,
  height   = 5,
  device   = "svg"
)
```



